Nonlinear Dendritic Coincidence Detection for Supervised Learning

### 2021-05-12 notes/frontiers_coincidence_learning/frontiers.tex
.

\def\keyFont{\fontsize{8}{11}\helveticabold }

@@

Cortical pyramidal neurons have a complex dendritic anatomy, whose
function is an active field of scientific research.
In particular, the segregation between its soma and the apical dendritic tree
is believed to play an active role in processing feed-forward 
sensory information and top-down or feedback signals. In this work, we use a simple 
two-compartment model accounting for the nonlinear interactions between
basal and apical input streams and show that a simple Hebbian learning rule
in the basal compartment allows the neuron to align its basal input to 
a target signal in the apical compartment. We show that this learning
process is robust against strong distractions in the basal input space
and demonstrate its effectiveness in a linear classification task.

@@

In recent years, a growing body of research has addressed the 
functional implications of the distinct physiology and anatomy of 
cortical pyramidal neurons \cite{Spruston2008,Hay2011,Ramaswamy2015}. 
In particular, on the theoretical side,
we saw a paradigm shift from treating neurons as point-like electrical
structures towards embracing the entire dendritic structure 
\cite{Larkum2009,Poirazi2009,Shai2015}. This was 
mostly due to the fact that experimental work uncovered dynamical properties
of these cells that simply could not be accounted for by point models
\cite{Spruston1995,Hausser2000}.

An important finding was that the apical dendritic tree of
cortical pyramidal neurons can act as a separate nonlinear synaptic 
integration zone \cite{Spruston2008,Branco2011}. 
Under certain conditions, a dendritic $\rm Ca^{2+}$ spike
can be elicited that propagates towards the soma, causing rapid, bursting
spiking activity. One of the cases in which dendritic spiking can occur
was termed `backpropagation-activated $\rm Ca^{2+}$ spike firing' 
(`BAC firing'): A single somatic spike can backpropagate towards the apical
spike initiation zone, in turn significantly facilitating the initiation of 
a dendritic spike \cite{Stuart2001,Spruston2008,Larkum2013}. 
This reciprocal coupling is believed to act as a form of
coincidence detection: If apical and basal synaptic input co-occurs, the 
neuron can respond with a rapid burst of spiking activity. 
The firing rate of these temporal bursts exceeds the firing 
rate that is maximally achievable under basal synaptic input alone, 
therefore representing a form of temporal coincidence
detection between apical and basal input.

Naturally, these mechanisms also affect plasticity and thus learning
within the cortex \cite{Sjoestroem2006,Ebner2019}. 
While the interplay between basal and apical stimulation and
its effect on synaptic efficacies is subject to ongoing research, 
there is some evidence that BAC-firing tends to shift plasticity 
towards long-term potentiation (LTP) \cite{Letzkus2006}. 
Thus, coincidence between basal and apical input appears 
to also gate synaptic plasticity.

In a supervised learning scheme, where the top down input
arriving at the apical compartment acts as the teaching signal,
the most straight-forward learning rule for the basal synaptic
weights would be derived from an appropriate loss function,
such as a mean square error, based on the difference between 
basal and apical input, i.e.\ $I_{\rm p} - I_{\rm d}$. Theoretical work has 
investigated possible learning mechanisms
that could utilize an intracellular error signal
\citep{Urbanczik2014,Schiess2016,Guerguiev2017}.
However, a clear experimental
evidence for a physical quantity encoding such an error 
is---to our knowledge---yet to be found. 
On the other hand, Hebbian-type plasticity is extensively
documented in experiments 
\cite{Gustafsson1987,Debanne1994,Markram1997,Bi1998}. 
Therefore, our work is based on the question whether the 
non-linear interactions between basal and apical synaptic input could, 
when combined with a Hebbian plasticity rule, allow a neuron
to learn to reproduce an apical teaching signal in its
proximal input.

In our work, we combine a phenomenological model that 
generates the output firing rate as a function of two 
streams of synaptic input (subsuming basal and apical input)
with Hebbian, as well as BCM-like plasticity rules on basal synapses. 
We hypothesized that this combination of neural 
activation and plasticity rules would lead to an
increased correlation between basal and apical input.

Furthermore, this temporal alignment could potentially 
facilitate apical inputs to act as top-down teaching signals, 
without the need for an explicit error-driven
learning rule. Thus, we also test our model in a simple linear 
supervised classification task and compare it with 
the performance of a simple point neuron equipped with similar 
plasticity rules.

@@

The neuron model used throughout this study is a discrete-time 
rate encoding model that uses two separate input variables, 
subsuming the total synaptic input current injected arriving 
at the basal (proximal) and apical (distal) 
dendritic structure of a pyramidal neuron, respectively. 
The model is a slightly simplified version of a phenomenological 
model proposed by \citet{Shai_2015}. Denoting the input currents 
$I_{\rm p}$ (proximal) and $I_{\rm d}$ (distal), 
the model is written as
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{align}
\begin{split}
y\left(t\right) &= \alpha  \sigma\left( I_{\rm p}(t) - \theta_{p0} \right)
\left[1-\sigma\left(I_{\rm d}(t) - \theta_d\right)\right] \\
&+ \sigma\left(I_{\rm d}(t) - \theta_d \right)
\sigma\left( I_{\rm p}(t) - \theta_{p1} \right)
\end{split} \label{eq:comp_model}\\
\sigma(x) &\equiv \frac{1}{1+\exp(-4x)} \; .
\end{align}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here, $\theta_{p0}$, $\theta_{p1}$ and $\theta_d$ are threshold 
variables with respect to proximal and distal input. 
Overall, this equation describes two distinct
regions of neural activation in the 
$(I_{\rm p}, I_{\rm d})$-space which differ in their
maximal firing rates, which are set to $1$ and 
$\alpha$, where $0 < \alpha < 1$.
A plot of \eqref{eq:comp_model} is shown in Fig.~\ref{fig:comp_model}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=0.6\columnwidth]{plot_comp_mod_marks.png}
\caption{{\bf Two-compartment rate model.} Firing rate as a
	function of proximal and distal input $I_{\rm p}$ and $I_{\rm d}$, 
	see \eqref{eq:comp_model}. 
	The thresholds $\theta_{p0}$, $\theta_{p1}$ and $\theta_d$ define
	two regions of neural activity with a a maximal firing rate of $1$
	and $\alpha$.}
\label{fig:comp_model}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In all our numerical experiments, we compare this model with a 
simple point neuron model that was given by
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
y(t) = \sigma\left(I_{\rm p}(t) + I_{\rm d}(t) - \theta \right) \; .
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The apical input $I_{\rm d}$ was generated
`as is', meaning, it is not dynamically calculated as a superposition
of multiple presynaptic inputs, but given by
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
I_{\rm d}(t) = n_d(t) x_d(t) - b_d(t) \; ,
\label{eq:I_d}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
where $n_d(t)$ is a scaling factor, $x_d(t)$ a pre-generated
discrete time sequence and $b_d(t)$ a bias. Note that $n_d$ and $b_d$ 
are time dependent since they are subject to adaptation processes 
that are described in the next section.

Similarly, $I_{\rm p}(t)$ is given by
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
I_{\rm p}(t) = n_p(t) \sum_{i=1}^{N} x_{p,i}(t) w_i(t) - b_p(t) \; ,
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
where $N$ is the number of presynaptic afferents, $x_{p,i}(t)$ the
corresponding sequences and $w_i(t)$ the synaptic efficacies.
As for $I_{\rm d}(t)$, $n_p(t)$ and $b_p(t)$ is a time dependent
scaling and bias.

%---------------------------------------%
\subsection{Plasticity}
%---------------------------------------%

We implemented a Hebbian plasticity rule for the basal 
synaptic weights given by the following update equation:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{align}
w_i(t+1) &= w_i(t) + \mu_w \left(x_{p,i}(t) - \tilde{x}_{p,i}(t)\right)
\left(y(t) - \tilde{y} \right)
\label{eq:hebb_plast} \\
\tilde{x}_{p,i}(t+1) &= (1-\mu_{av})\tilde{x}_{p,i}(t) + \mu_{av}x_{p,i}(t+1) \\
\tilde{y}(t+1) &= (1-\mu_{av})\tilde{y}(t) + \mu_{av}y(t+1)
\end{align}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Additionally, we use a
synaptic normalization constraint
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
w_i(t) \rightarrow \frac{w_i(t)}{\lVert \mathbf{w}(t)\rVert}
\label{eq:weight_norm}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
in each time step, where $\lVert \mathbf{w}(t)\rVert$ denotes the Euclidean
norm of the synaptic weight vector.

For comparative reasons, the point neuron model is equipped with the
same plasticity rule for the proximal weights as \eqref{eq:hebb_plast}.

Additionally, the scaling and bias variables are changing dynamically
according to the following homeostatic plasticity rules:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{align}
b_p(t+1) &= b_p(t) + \mu_b \left[I_{\rm p}(t) - I_{\rm p}^t\right] \\
b_d(t+1) &= b_d(t) + \mu_b \left[I_{\rm d}(t) - I_{\rm d}^t\right] \\
n_p(t+1) &= n_p(t) + 
\mu_n \left[V_p^t - \left( I_{\rm p}(t) - \tilde{I}_p(t)\right)^2\right] \\
n_d(t+1) &= n_d(t) + 
\mu_n \left[V_d^t - \left( I_{\rm d}(t) - \tilde{I}_d(t)\right)^2\right] \\
\tilde{I}_p(t+1) &= (1-\mu_{av})\tilde{I}_p(t) + \mu_{av} I_{\rm p}(t+1) \\
\tilde{I}_d(t+1) &= (1-\mu_{av})\tilde{I}_d(t) + \mu_{av} I_{\rm d}(t+1)
\end{align}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here, $I_{\rm p}^t$, $I_{\rm d}^t$, $V_p^t$ and $V_p^t$ define targets for the 
temporal means and variances of $I_{\rm p}$ and $I_{\rm d}$. 
The dynamic variables $\tilde{I}_p$ and $\tilde{I}_d$ are simply 
low-pass filtered running averages of $I_{\rm p}$ and $I_{\rm d}$.

A list of all parameter values is given in Table~\ref{tab:parameters}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
	\begin{tabular}{ l | l || l | l }
		$\theta_{p0}$ & $0$ & $V_d^t$ & $0.25$ \\
		$\theta_{p1}$ & $-1$ & $\mu_b$ & $10^{-3}$ \\ 
		$\theta_{d}$ & $0$ & $\mu_n$ & $10^{-4}$ \\  
		$\alpha$ & $0.3$ & $\mu_{av}$ & $5 \cdot 10^{-3}$ \\   
		$\mu_w$ & $5 \cdot 10^{-5}$ & $I_{\rm p}^t$ & $0$ \\
		$V_p^t$ & $0.25$ & $I_{\rm d}^t$ & $0$  
	\end{tabular}
	\caption{Model parameters}
	\label{tab:parameters}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%---------------------------------------%
\subsection{Increased Alignment between Basal and Apical Inputs}
\label{sect:alignment}
%---------------------------------------%

As a first test, we wanted to quantify the neuron's ability to 
align its basal input to the apical teaching signal.
To do so, we defined the pearson correlation coefficient
$\rho[I_{\rm p},I_{\rm d}]$ between the basal and apical input current
as our measure of interest. We determined this temporal
correlation coefficient after simulating all plasticity mechanisms
under stationary random input sequences with certain statistical
properties that shall be explained in the following.
As a starting point, we chose all $x_{p,i}(t)$ to be randomly
drawn from a uniform distribution, where $x_{p,i}(t) \in (0,1)$.
For $I_{\rm d}(t)$ to be fully `reconstructable' by the basal input, 
$x_d(t)$ had to be some linear combination of all $x_{p,i}(t)$.
Therefore, we chose $x_d(t) = \sum_{i=1}^N a_i x_{p,i}(t)$, where
$\mathbf{a}$ is a random vector with unit length.
Since we used a Hebbian learning scheme, we expected that 
the direction and magnitude of the principal components of 
the basal input would also significantly affect the outcome of
the experiment: A large variance in the basal input 
orthogonal to the `reconstruction vector' $\mathbf{a}$ 
should act as a distraction for the plasticity and reduce the
resulting alignment between $I_{\rm p}$ and $I_{\rm d}$. Therefore, we 
applied a transformation to the input sequences $x_{p,i}(t)$
that were parameterized by two quantities, a scaling factor $s$
and the dimension $N_{dist}$ of a subspace of the basal input space.
A set of $N_{dist}$ orthonormal basis vectors was randomly generated,
which were also orthogonal to $\mathbf{a}$. 
Withing this $N_{dist}$-dimensional subspace, the input sequences 
$x_{p,i}(t)$ were then rescaled by the factor $s$.

After both $x_{p,i}(t)$ and $x_d(t)$ were generated,
a simulation was run using all previously described plasticity
mechanisms until the dynamic variables reached a stationary state.
After this learning phase, another set of input sequences was generated
using the previously described protocol and $\rho[I_{\rm p},I_{\rm d}]$ 
was calculated. Note that plasticity was turned off in this phase. 
This entire procedure allowed us to calculate $\rho[I_{\rm p},I_{\rm d}]$ 
as a function of the distraction parameters $s$ and $N_{dist}$. 
The results for both neuron models is shown in 
Fig.~\ref{fig:corr_dimension_scaling}. Here, the total number of 
basal inputs is $N=10$. One can observe a decorrelation transition for 
both models. However, the compartment model supports a 
significantly stronger distraction in terms of the scaling 
factor $s$ as compared to the point model. This is a first confirmation 
of our hypothesis that nonlinear interactions between basal and 
apical input could improve learning guided by top-down signals.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=0.7\columnwidth]{corr_dimension_scaling_high_input_dim}
\caption{{\bf Alignment between Basal and Apical Input.} Color
	encodes the Pearson correlation $\rho[I_p,I_d]$ for different
	number of orthogonal distraction directions $N_{dist}$ 
	and the corresponding scaling faction $s$.}
\label{fig:corr_dimension_scaling}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\def\keyFont{\fontsize{8}{11}\helveticabold }
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Cortical pyramidal neurons have a complex 
dendritic anatomy, whose function is an active 
field of scientific research. In particular, 
the segregation between its soma and the apical 
dendritic tree is believed to play an active 
role in processing feed-forward sensory 
information and top-down or feedback signals. 
In this work, we use a simple two-compartment 
model accounting for the nonlinear interactions 
between basal and apical input streams and show 
that a simple Hebbian learning rule in the basal 
compartment allows the neuron to align its basal 
input to a target signal in the apical compartment. 
We show that this learning process, termed 
coincidence detection, is robust against strong 
distractions in the basal input space and 
demonstrate its effectiveness in a linear 
classification task.

@@

In recent years, a growing body of research has addressed the 
functional implications of the distinct physiology and anatomy of 
cortical pyramidal neurons \citep{Spruston2008,Hay2011,Ramaswamy2015}. 
In particular, on the theoretical side,
we saw a paradigm shift from treating neurons as point-like electrical
structures towards embracing the entire dendritic structure 
\citep{Larkum2009,Poirazi2009,Shai2015}. This was 
mostly due to the fact that experimental work 
uncovered dynamical properties of pyramidal neuronal
cells that simply could not be accounted for by point models
\citep{Spruston1995,Hausser2000}.

An important finding is that the apical dendritic tree of
cortical pyramidal neurons can act as a separate nonlinear synaptic 
integration zone \citep{Spruston2008,Branco2011}. 
Under certain conditions, a dendritic $\rm Ca^{2+}$ spike
can be elicited that propagates towards the soma, causing rapid, bursting
spiking activity. One of the cases in which dendritic spiking can occur
was termed `backpropagation-activated $\rm Ca^{2+}$ spike firing' 
(`BAC firing'): A single somatic spike can backpropagate towards the apical
spike initiation zone, in turn significantly facilitating the initiation of 
a dendritic spike \citep{Stuart2001,Spruston2008,Larkum2013}. 
This reciprocal coupling is believed to act as a form of
coincidence detection: If apical and basal synaptic input co-occurs, the 
neuron can respond with a rapid burst of spiking activity. 
The firing rate of these temporal bursts exceeds the firing 
rate that is maximally achievable under basal synaptic input alone, 
therefore representing a form of temporal coincidence
detection between apical and basal input.

Naturally, these mechanisms also affect plasticity and thus learning
within the cortex \citep{Sjoestroem2006,Ebner2019}. 
While the interplay between basal and apical stimulation and
its effect on synaptic efficacies is subject to ongoing research, 
there is evidence that BAC-firing tends to shift plasticity 
towards long-term potentiation (LTP) \citep{Letzkus2006}. 
Thus, coincidence between basal and apical input appears 
to also gate synaptic plasticity.

In a supervised learning scheme, where the top down input
arriving at the apical compartment acts as the teaching signal,
the most straight-forward learning rule for the basal synaptic
weights would be derived from an appropriate loss function,
such as a mean square error, based on the difference between 
basal and apical input, i.e.\ $I_p - I_d$,
where indices $p$ and $d$ denote `proximal' and
`distal', in equivalence to basal and apical. 
Theoretical studies have investigated possible 
learning mechanisms that could utilize an 
intracellular error signal
\citep{Urbanczik2014,Schiess2016,Guerguiev2017}.
However, a clear experimental
evidence for a physical quantity encoding such an error 
is---to our knowledge---yet to be found. 
On the other hand, Hebbian-type plasticity is extensively
documented in experiments 
\citep{Gustafsson1987,Debanne1994,Markram1997,Bi1998}. 
Therefore, our work is based on the question whether the 
non-linear interactions between basal and apical synaptic input could, 
when combined with a Hebbian plasticity rule, allow a neuron
to learn to reproduce an apical teaching signal in its
proximal input.

We investigate coincidence learning by
combining a phenomenological model that 
generates the output firing rate as a function 
of two streams of synaptic input (subsuming basal 
and apical inputs) with Hebbian, as well as 
BCM-like plasticity rules on basal synapses. 
In particular we hypothesized that this combination of neural 
activation and plasticity rules would lead to an
increased correlation between basal and apical inputs.
Furthermore, the temporal alignment observed in our study 
could potentially facilitate apical inputs to act as 
top-down teaching signals, without the need for an 
explicit error-driven learning rule. Thus, we also 
test our model in a simple linear supervised 
classification task and compare it with the 
performance of a simple point neuron equipped with 
similar plasticity rules.

@@

The neuron model used throughout this study 
is a discrete-time rate encoding model that 
contains two separate input variables, 
subsuming the total synaptic input current injected arriving 
at the basal (proximal) and apical (distal) 
dendritic structure of a pyramidal neuron, respectively. 
The model is a slightly simplified version of a phenomenological 
model proposed by \citet{Shai_2015}. Denoting the input currents 
$I_p$ (proximal) and $I_d$ (distal), 
the model is written as
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{align}
\begin{split}
y\left(t\right) &= \alpha  \sigma\left( I_p(t) - \theta_{p0} \right)
\left[1-\sigma\left(I_d(t) - \theta_d\right)\right] \\
&+ \sigma\left(I_d(t) - \theta_d \right)
\sigma\left( I_p(t) - \theta_{p1} \right)
\end{split} 
\label{eq_comp_model}\\
\sigma(x) &\equiv \frac{1}{1+\exp(-4x)} \; .
\end{align}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here, $\theta_{p0}>\theta_{p1}$ and $\theta_d$ 
are threshold variables with respect to proximal 
and distal inputs. Overall, equation 
(\ref{eq_comp_model}) describes two distinct
regions of neural activation in the 
$(I_p, I_d)$-space which differ in their
maximal firing rates, which are set to $1$ and 
$\alpha$, where $0 < \alpha < 1$.
A plot of \eqref{eq_comp_model} is shown 
in Fig.~\ref{fig:comp_model}.

When both input currents $I_d$ and $I_p$ 
are large, viz larger than the 
thresholds $\theta_d$ and $\theta_{p1}$,
the second term in \eqref{eq_comp_model}
dominates, which leads to $y\approx 1$. 
An intermediate activity plateau, of
strength $\alpha$ emerges in addition 
when $I_p>\theta_{p0}$ and 
$I_d<\theta_{d}$. As such, the compartment
model \eqref{eq_comp_model} is able to
distinguish neurons with a normal activity 
level, here encoded by $\alpha=0.3$, and
strongly bursting neurons, where the maximal
firing rate is unity. The intermediate plateau
allows neurons to process the proximal 
inputs $I_p$ even in the absence of distal
stimulation. The distal current $I_d$
acts therefore as an additional modulator.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\centering
\includegraphics[width=0.6\columnwidth]{plot_comp_mod_marks.png}
\caption{{\bf Two-compartment rate model.} 
The firing rate as a function of proximal and distal 
inputs $I_p$ and $I_d$, 
see \eqref{eq_comp_model}. 
The thresholds $\theta_{p0}$, $\theta_{p1}$ and 
$\theta_d$ define two regions of neural activity,
with a a maximal firing rate of unity an a plateau
at $\alpha=0.3$.}
\label{fig:comp_model}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In our numerical experiments we compare 
the compartment model with a classical point 
neuron, as given by
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
y(t) = \sigma\left(I_p(t) + I_d(t) - 
\theta \right) \; .
\label{eq_point_neuron}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The apical input $I_d$ is generated
`as is', meaning, it is not dynamically 
calculated as a superposition of multiple 
presynaptic inputs. To be concrete, we
used
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
I_d(t) = n_d(t) x_d(t) - b_d(t) \; ,
\label{eq_I_d}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
where $n_d(t)$ is a scaling factor, $x_d(t)$ 
a pre-generated discrete time sequence and 
$b_d(t)$ a bias. Note that $n_d$ and $b_d$ 
are time dependent since they are subject 
to adaptation processes, which will be
described in the next section. Similarly, 
the proximal input $I_p(t)$ is given by
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
I_p(t) = n_p(t) \sum_{i=1}^{N} 
x_{p,i}(t) w_i(t) - b_p(t) \; ,
\label{eq_I_p}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
where $N$ is the number of presynaptic afferents, 
$x_{p,i}(t)$ the corresponding sequences, 
$w_i(t)$ the synaptic efficacies and
$n_p(t)$ and $b_p(t)$ the (time dependent)
scaling and bias. Tyical values for the 
parameters used throughout this study
are presented Table~\ref{tab_parameters}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[b]
\caption{Model parameters,
as defined in sections \ref{sect:neuronmodel} and
\ref{sect_plasticity}.}
\begin{tabular}{ l | l || l | l }
		$\theta_{p0}$ & $0$ & $V_d^t$ & $0.25$ \\
		$\theta_{p1}$ & $-1$ & $\mu_b$ & $10^{-3}$ \\ 
		$\theta_{d}$ & $0$ & $\mu_n$ & $10^{-4}$ \\  
		$\alpha$ & $0.3$ & $\mu_{\rm av}$ & $5 \cdot 10^{-3}$ \\   
		$\mu_w$ & $5 \cdot 10^{-5}\quad$ & $I_p^t$ & $0$ \\
		$V_p^t$ & $0.25$ & $I_d^t$ & $0$  
\end{tabular}
\label{tab_parameters}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%---------------------------------------%
\subsection{Plasticity\label{sect_plasticity}}
%---------------------------------------%

We implemented a Hebbian plasticity rule for 
the proximal synaptic weights by the following 
update equation:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{align}
w_i(t+1) &= w_i(t) + \mu_w \left(x_{p,i}(t) - \tilde{x}_{p,i}(t)\right)
\left(y(t) - \tilde{y} \right)
\label{eq_hebb_plast} \\
\tilde{x}_{p,i}(t+1) &= 
(1-\mu_{\rm av})\tilde{x}_{p,i}(t) + \mu_{\rm av}x_{p,i}(t) 
\label{eq_tilde_x_p_i}\\
\tilde{y}(t+1) &= (1-\mu_{\rm av})\tilde{y}(t) + \mu_{\rm av}y(t)
\label{eq_tilde_y}
\end{align}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The trailing time averages $\tilde{x}_{p,i}$ and
$\tilde{y}$, respectively of the presynaptic 
basal activites, $x_{p,i}$, and of the neural
firing rate $y$, enter the Hebbian learing
rule (\ref{eq_hebb_plast}) as reference levels.
Pre- and post-synaptic neurons are considered
to be active/inactive when being above/below
the respective trailing averages. The
timescale of the averaging, $1/\mu_{\rm av}$,
is typcially over 200 time steps, see
Table~\ref{tab_parameters}. 
With $1/\mu_w=2\cdot10^4$ learning is 
assumed to be be considerably slower, 
as usual for statistical update rules.
Additionally, 
we use a synaptic normalization constraint,
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
w_i(t) \rightarrow \frac{w_i(t)}{\lVert 
\mathbf{w}(t)\rVert}\;,
\label{eq:weight_norm}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
in each time step, where 
$\lVert \mathbf{w}(t)\rVert$ denotes the Euclidean
norm of the synaptic weight vector. For comparative 
reasons, the point neuron model is equipped with the
same plasticity rule for the proximal weights as 
(\ref{eq_hebb_plast}).

The bias variables entering the definitions
\eqref{eq_I_d} and \eqref{eq_I_p} of the distal
proximal currrent, $I_d$ and $I_p$, 
are assumed to adapt accodingly to
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{align}
\label{eq_b_p_dot}
b_p(t+1) &= b_p(t) + \mu_b \left[I_p(t) - I_p^t\right] \\
b_d(t+1) &= b_d(t) + \mu_b \left[I_d(t) - I_d^t\right] \;,
\label{eq_b_d_dot}
\end{align}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
where $I_p^t=0$ and, $I_d^t=0$
are preset targets and $1/\mu_b=10^3$ the
timescale for the adaption process. Over
time, both the distal and the proximal 
currents, $I_d$ and $I_p$, average out.

Adaption rules for the bias entering a 
transfer function, such as
\eqref{eq_b_d_dot} and \eqref{eq_b_p_dot},
have the task to regulate overall activitiy
levels. The overall magnitude of the
synaptic weights, which are determined by 
synaptic rescaling factors, here $n_d$ and $n_p$, 
as defined in \eqref{eq_I_d} and \eqref{eq_I_p},
will regulate in contrast the variance of
the neural activity, and not the average
level \citep{schubert2021local}. In this
spirit we consider
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{align}
n_d(t+1) &= n_d(t) + 
\mu_n \left[V_d^t - \left( I_d(t) - \tilde{I}_d(t)\right)^2\right] \\
n_p(t+1) &= n_p(t) + 
\mu_n \left[V_p^t - \left( I_p(t) - \tilde{I}_p(t)\right)^2\right] \\
\tilde{I}_d(t+1) &= (1-\mu_{\rm av})\tilde{I}_d(t) + \mu_{\rm av} I_d(t) \\
\tilde{I}_p(t+1) &= (1-\mu_{\rm av})\tilde{I}_p(t) + \mu_{\rm av} I_p(t) 
\end{align}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here, $V_p^t$ and $V_p^t$ define targets for 
the temporal averaged variances of $I_p$ 
and $I_d$.  
The dynamic variables $\tilde{I}_p$ and $\tilde{I}_d$ 
are simply low-pass filtered running averages of 
$I_p$ and $I_d$. Overall, the framework
specified here allows the neuron to be fully flexible,
as long as the activtiy level and its variance 
fluctuate around preset target values 
\citep{schubert2021local}. A list of 
the parameter values used throughout this investigation 
is given in Table~\ref{tab_parameters}. Our choices of
target means and variances are based on the assumption
that neural input should be tuned towards a certain
working regime of the neural transfer function. In
the case of the presented model, this means that both
proximal and distal input cover an area where the nonlinearities
of the transfer function are reflected without oversaturation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\centering
\includegraphics[width=0.55\columnwidth]{illustration_classification}
\caption{{\bf Input Space for the Linear Classification Task.}
Two clusters of presynaptic basal activities were generated from 
multivariate Gaussian distributions. Here, $s$ denotes the standard
deviation orthogonal to the normal vector $\mathbf{a}$ of the 
classification hyperplane, as defined by \eqref{eq_x_d_a}.}
\label{fig_illustration_classification}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%---------------------------------------%
\subsection{Alignment between Basal and Apical Inputs}
\label{sect:alignment}
%---------------------------------------%

As a first test, we quantify the neuron's ability to 
align its basal input to the apical teaching signal.
This can be done using the pearson correlation coefficient
$\rho[I_p,I_d]$ between the basal and 
apical input currents. We determined 
$\rho[I_p,I_d]$ after the simulation,
which involves all plasticity mechanisms, both
for the synaptic weights and for the intrinsic 
parameters. The input sequences $x_{p,i}(t)$ 
is randomly drawn from a uniform distribution,
in $[0,1]$, which is done independently for 
each $i\in[1,N]$.

For the distal current $I_d(t)$ to be fully 
`reconstructable' by the basal input, $x_d(t)$ has 
to be a linear combination 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{align}
x_d(t) &= \sum_{i=1}^N a_i x_{p,i}(t)
\label{eq_x_d_a}
\end{align}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
of the $x_{p,i}(t)$, where the $a_i$ are the
components of a random vector $\mathbf{a}$ 
of unit length.

Given that we use with \eqref{eq_hebb_plast}
a Hebbian learning scheme, one can expect that
the direction and the magnitude of the principal 
components of the basal input may affect the
outcome of the simulation significantly:
A large variance in the basal input 
orthogonal to the `reconstruction vector' $\mathbf{a}$ 
is a distraction for the plasticity. The
observed temporal alignment between $I_p$ and 
$I_d$ should hence suffer when such a 
distraction is present. The situaton is illustrated 
in Fig.~\ref{fig_illustration_classification}.

In order to test the effects of distracting directions,
we applied a transformation to the input sequences $x_{p,i}(t)$.
For the transformation two parameters are used,
a scaling factor $s$ and the dimension $N_{\rm dist}$ 
of the distracting subspace within the basal input 
space. The $N_{\rm dist}$ randomly generated
basis vectors are orthogonal to the superposition
vector $\mathbf{a}$, as defined by \eqref{eq_x_d_a},
and to each others. Within this $N_{\rm dist}$-dimensional 
subspace, the input sequences $x_{p,i}(t)$ are
rescaled subsequently by the factor $s$. 
See Fig.~\ref{fig_illustration_classification}.
After the learning phase, a second set of input 
sequences $x_{p,i}(t)$ and $x_d(t)$ is generated for 
testing purposes, using the identical protocol, and
the cross correlation $\rho[I_p,I_d]$ 
evaluated. During the testing phase plasticity is 
turned off.

The overall aim of our portocal is to evaluate
the degree $\rho[I_p,I_d]$ to which
the proximal current $I_p$ aligns in the
temporal domain to the distal input $I_d$.
We recall that this is a highly non-trivial
question, given that the proximal synpatic
weights are adapted via Hebbian plasticity, 
see \eqref{eq_hebb_plast}. The error 
$(I_p-I_d)^2$ does not
enter the adaption rules employed.
Results are presented in
Fig.~\ref{fig_corr_dimension_scaling}
as a function of the distraction parameters 
$s$ and $N_{\rm dist}\in[0,N-1]$. The
total number of basal inputs is $N=100$.

For a comparison, in 
Fig.~\ref{fig_corr_dimension_scaling}
data for both the compartment model
and for a point neuron are presented,
as defined respectively by
\eqref{eq_comp_model}
and \eqref{eq_point_neuron}.
One can observe a decorrelation transition for 
both models. However, the compartment model supports a 
significantly stronger distraction in terms of the scaling 
factor $s$ as compared to the point model. This is a first confirmation 
of our hypothesis that nonlinear interactions between basal and 
apical input could improve learning guided by top-down signals.
